---
title: "Homework 3"
output:
  html_document: default
  pdf_document: default
date: 'Deadline: 9 February, 12:00'
subtitle: 'Linguistic Data: Quantitative Analysis and Visualisation'
---

The solutions should be submitted via GitHub.  

## Part 1. A preliminary training  

*Do not use R (RStudio) to solve problems in Part 1. Answers won't be evaluated. *

### Problem 1

Look at the following histogram and answer the questions.

```{r, echo=FALSE}
set.seed(1234)
x <- rnorm(20, mean=3, sd=6)
hist(x, col="tomato", main="")
```

a. What is the proportion of values in the sample that exceed $5$? Explain your answer.
1/4x + 10x + 4x doesn't exceed $5$ and 4x + 1/4x exceed $5$, so the proportion is 17/4 * 4/57 ~ 0.298

b. Indicate the interval where the median of this sample can lie. Explain your answer.
The median's interval is from -5 to 0, because most of x lies in this interval.

c. How the histogram will change if we add an element $7$ to the sample? Explain your answer.
The frequency of the interval 5-10 will increase a little bit, other frequences will remain the same, because there will be 1 more element in the interval 5-10 which means that frequency will increase.

### Problem 2

a. Look at the histograms of two samples. They illustrate the distribution of normalized average reaction time to frequent words (in ms) in two groups of people.

```{r, echo=FALSE, fig.height=3}
set.seed(1234)
sample1 <- rnorm(100, 500, 50)
sample2 <- rnorm(100, 500, 20)
hist(sample1, col="lightblue", xlim=c(100, 900))
```

```{r, echo=FALSE, fig.height=3}
hist(sample2, col="lightgreen", xlim=c(100, 900))
```

Which of the samples has a larger variance? Explain your answer.
The first sample has the larger variance, because ~350 - 650 is a larger interval then ~450 - 550.

\newpage

b. Look at the histograms of two samples.

```{r, echo=FALSE, fig.height=8}
set.seed(12)
sample1 <- rnorm(100, 3, 5)
sample2 <- rnorm(100, 10, 0.8)
par(mfrow=c(2,2))
hist(sample1, col="lightblue", xlim=c(-40, 40))
hist(sample2, col="lightgreen", xlim=c(5, 15))
```

Which of the samples has a larger variance? Explain your answer.
The first one, because ~-10 - 15 is a larger interval then 8 - 12.

## Part 2

*Do not use R (RStudio) to solve problems in Part 2. Answers for problem 3 will be evaluated. Please paste YES or NO into (empty) code blocks and explain you answer below the block.*

### Problem 3

Below is the histogram of the number of mistakes students made while writing an examination essay in English. Look at the histogram and answer the questions.

```{r, echo=FALSE, fig.height=4}
set.seed(2)
mistakes <- rbeta(1000, 2, 7)*100
hist(mistakes, breaks=60, col= "deepskyblue", ylim=c(0, 40))
```

### 3.1
Is it true that $50$% students made more than $35$ mistakes?
```
NO
```
Explain your answer below:
A part of the figure to the left of 35 mistakes has much more samples then the right part.

### 3.2
Is it true that most students made no more than $10$ mistakes?
```
NO
```
Explain your answer below:
There are more students that made 10-80 mistakers then 0-10.

### 3.3
Which of the following values is closer to be the median of `mistakes`: $10$, $20$, $30$, $40$?
```
$10$
```
Explain your answer below:
The hughest frequences seem to be around $10$ mistakes.

### Problem 4. Exact binomial test

The null hypothesis is that $p=0$ (i.e. no success is possible). In a dataset, there is only one success out of 1 000 000 observations. Will you reject the null hypothesis? 
```
YES
```
Explain your answer below:
p-value is 1/1 000 000 which is less then 0.05.

## Part 3

*Use R (RStudio) to solve problems in Part 3. Your answers will be evaluated. Please paste R code into R code blocks and explain you answer below the block, if needed. *

### Problem 5 

Here is a sample of respondents' age:  

$44$, $50$, $42$, $64$, $66$, $42$, $72$, $56$, $72$, $54$, $46$, $48$, $48$, $52$, $50$, $66$, $84$.

### 5.1
Arrange them in a vector and call it `age`. 
```{r}
age <- c(44, 50, 42, 64, 66, 42, 72, 56, 72, 54, 46, 48, 48, 52, 50, 66, 84)
```

### 5.2
Examine the type of `age` variable (numeric, character, etc).
```
class(age)
```

### 5.3
Plot the histogram of the vector `age` with $5$ bins. Change its color to any you want. (Use either R basic or ggplot2 style for plotting.)
```{r}
hist(age, col='lightyellow', breaks = 4)
```


### Problem 6

Here is a series of words:  
*pie, bar, bar, pie, pie, bar, bar, chart*.

### 6.1
Arrange elements above in a vector and call it `words` 
```{r}
words <- c("pie", "bar", "bar", "pie", "pie", "bar", "bar", "chart")
```

### 6.2
Calculate the relative frequences of values in `words` measured in percent.
```{r}
table(words)/length(words)
```

## Problem 7. Position of verbs in verses

The dataset [“The last words in verses”](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv) is based on texts written in the 1820s and 1920s (Corpus of Russian Poetry of the Russian National Corpus). Authors collected only one line per author to keep observations as independent from each other as possible. 

Variables:  
- Decade — decade of creation: 1820s, 1920s.  
- RhymedNwords — the number of words in the rhyming position (usually one word, but there are two words in cases such as _вина бы_ 'I would like to get) wine' (which is rhymed with _жабы_ 'toad', see http://russian-poetry.ru/Poem.php?PoemId=18261)).  
- RhymedNsyl — the number of syllables in the rhyming position.  
- UPoS — part of speech of the last word. 
- LineText — a sampled verse.  
- Author — the author of the text.  

Can we decide that in verses written in 1920s, verbs in the rhyming posision are used differently (more often or less often) than expected for verbs in general? 

Let's assume that the probability for verbs to be used in any position ('in general') is 17% (according to [](http://www.ruscorpora.ru/new/corpora-stat.html) ).

### 7.1 State hypothesis

What is your null hypothesis $H_0$ and what is the alternative hypothesis $H_1$?
```
H0: Probability for the verbs to be used in verses written in 1920s equals the probability for verbs to be used in any position equals 17%.
H1: Probability for the verbs to be used in verses written in 1920s is not equal to the probability for verbs to be used in any position.
```

### 7.2

Read the dataset [“The last words in verses”](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv). Filter out the relevant observations from 1920s, calculate the number of verbs observed in the sample, and the sample size. 
```{r}
df <- read.csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv", sep = "\t")
year9 <- df[df$Decade=='1920s',]
total <- nrow(year9) #the sample size
verbs <- year9[year9$UPoS=="VERB",]
total_verbs <- nrow(verbs) #the number of verbs observed in the sample
total_verbs
total
```

### 7.3

Use an exact binomial test to calculate p-value. 
```{r}
binom.test(total_verbs, total, p=0.17)
```

### 7.4 Interpret results

Give your interpretation of obtained p-value. Answer the initial question: Can we decide that in verses written in 1920s, verbs are used in the rhyming posision more often or less often than expected?
```
True probability of success is not equal to 0.17, which means that probability for the verbs to be used in verses written in 1920s is not equal to the probability for verbs to be used in any position. 

In verses written in 1920s, verbs are used in the rhyming posision less often than expected.
```

### 7.5 
*(A bonus problem, extra points in evaluation)*.
Repeat 2.3 for verses written in the 1820s.
```{r}
year8 <- df[df$Decade=='1820s',]
total <- nrow(year8) #the sample size
verbs <- year8[year8$UPoS=="VERB",]
total_verbs <- nrow(verbs) #the number of verbs observed in the sample
total_verbs
total

binom.test(total_verbs, total, p=0.17)
```

Write down your general conlusions about data provided for both 1920s and 1820s data.
```
In verses written in 1920s, verbs are used in the rhyming posision less often then in verses written in 1820s.
```

### Problem 8. One-sample t-test
Using Icelandic data on vowel duration from seminar [Link](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv) test the null hypothesis that the population mean of vowel duration in speaker `shg05` equals 73 (ms).
To perform a one-sample t-test, you can use the following example of R code: 
```
t.test(sample, mu = 7725) # mu is a population mean
```

### 8.1
Write down a two-tailed alternative hypothesis.
```
H1: The population mean of vowel duration in speaker `shg05` is either less then 73 or more then 73.
```

### 8.2
Perform a one-sample t-test. 
```{r}
phono <- read.csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv")

sample <- phono[phono$speaker=="shg05", "vowel.dur"]
t.test(sample, mu = 73) 
```

### 8.3
Interpret results. 
```
The population mean of vowel duration in speaker `shg05` equals 70 ms.
```

#### Supplementary reading
Use of exact binomial test in lingiistic research:  

* Gries, Stefan Th. "Phonological similarity in multi-word units." Cognitive Linguistics 22.3 (2011): 491-510. [Link](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.186.7412&rep=rep1&type=pdf)  
Stefan Gries proves that alliteration is observed in multi-word expressions more often than in general.

* Harald Bayen (2008: 51-52) evaluates the probability of observing exactly one occurrence of the word _hare_ in the corpus sample of 1 mln words given its estimated frequency of 8.23 words per million according to the SELEX frequency database.

On measures of central tendency: 

* Levshina 2015, Chapter 3 (p. 48); Gries 2009, Chapter 1.3 (p. 116). 

On t-test: 

* Gries 2009, Chapter 3 (p. 198). 